---
title: "idr_filt"
author: "Qian Hui TAN"
date: "`r Sys.Date()`"
format: gfm
toc: true
number-sections: true
---
In this notebook, we filter the IDR peaks to only contain high-quality 
peaks (IDR 0 - 0.05).

# Setup

## Load libraries

```{r}
suppressPackageStartupMessages({
    # Load libraries
    library(ChIPseeker)
    library(GenomicRanges)
})
```


## Load data

```{r}
deaf1_bed <- readPeakFile("deaf1_idr_all.bed")
```

```{r}
length(deaf1_bed)
```

# Data wrangling


`deaf1_bed` has a few columns that are named V4 to V14. We rename them so they make sense:

```{r}
head(deaf1_bed)
```

We rename the metadata columns (mcols) with the correct names: 

```{r}
colnames(mcols(deaf1_bed)) <- c(
                  "idr_name", "scaled_idr", "idr_strand",
                   "local_idr", "global_idr",
                   "rep1_chromstart", "rep1_chromend", "rep1_signal",
                   "rep2_chromstart", "rep2_chromend", "rep2_signal"
                  )
```

```{r}
head(deaf1_bed)
```

While this IDR file has IDR values, it has not been filtered. We know this because we should only have 608/1550 peaks if we check the IDR output, but here we have 1550 peaks. 

To be sure, we do a quick summary of the scaled IDR values, min(int(log2(-125IDR), 1000). Peaks with an IDR of 0 have a score of 1000, idr 0.05 have a score of int(-125log2(0.05)) = 540, and idr 1.0 has a score of 0.

```{r}
summary(deaf1_bed$scaled_idr)
```

Min score is 121 instead of 540 - that means they have not been filtered out. We filter them out: 

```{r}
# Filter and keep only highly reproducible peaks:
# IDR 0 - 0.05 (scaled IDR 540 - 1000)
deaf1_filt <- deaf1_bed[deaf1_bed$scaled_idr >= 540, ]
length(deaf1_filt)

head(deaf1_filt)
```

That's strange. Why do we have 608 peaks here, but 607 from the cutoff? 

Let's take a look at the peaks that barely pass the cutoff: 

```{r}
deaf1_bed[deaf1_bed$scaled_idr >= 540 & deaf1_bed$scaled_idr < 542,]
```
We have 2 peaks with scaled_idr 540, and another 1 with scaled_idr = 541. Since they are `integers`, I'm guessing there was a rounding error. 

In any case, we include both the peaks with scaled_idr = 540. 

# Exporting 

```{r}
write.table(deaf1_filt, "deaf1_idr_filt.bed", quote = F, 
            col.names = FALSE, row.names = FALSE, sep = "\t")

saveRDS(deaf1_filt, "deaf1_idr_filt.RDS")
```


```{r}
sessionInfo()
```

